---
title: "TARASPINA VP | 16S miTags | Eukaryota"
subtitle: "Dataset: mapping_SILVA_plus_PhytoRef_plus_CyanoDB_NR99.txt"
date: "27.01.2015"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    highlight: zenburn
    toc: yes
---

<!--- INITIALIZATION
```{r, echo=FALSE}
#error hook to kill knitr in case of errors
library(knitr)
knit_hooks$set(error = function(x, options) stop(x))
opts_chunk$set(cache=TRUE, autodep=TRUE)
```
--->

# 1) Data overview

Samples with less than 167 reads are removed: 

``` {r load_data, echo=FALSE, message=FALSE}
#setwd("/genwork/lrubinat")
setwd("/home/laura/Documents/TFM/genwork/data_analysis/TP_VP_16S_miTags_99tb/TP_VP_16S_miTags_99tb_euk/")

#read data 
tp_vp_16SmiTags_e <- read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/TP_VP_16S_miTags_99tb_euk.txt", head=TRUE, fill=TRUE)

#table dimensions and format before setting column names
#dim(tp_vp_16SmiTags_e) # 7206  274
#tp_vp_16SmiTags_e[1:5,1:5]

#row names = OTU name (option A)
row.names(tp_vp_16SmiTags_e)<-tp_vp_16SmiTags_e[,1]

tp_vp_16SmiTags_e<-tp_vp_16SmiTags_e[,-1]
tp_vp_16SmiTags_e[is.na(tp_vp_16SmiTags_e)]<-0

dim(tp_vp_16SmiTags_e)
tp_vp_16SmiTags_e[1:5,1:5]

#remove samples with less than 167 reads
tp_vp_16SmiTags_e_min167<-tp_vp_16SmiTags_e[,colSums(tp_vp_16SmiTags_e) > 167]
amplicons_per_sample_tp_vp_16SmiTags_e<-colSums(tp_vp_16SmiTags_e)
amplicons_per_sample_tp_vp_16SmiTags_e[which(colSums(tp_vp_16SmiTags_e)<167)]
#there are no samples with less than 167 reads

#amplicons_per_sample_tp_vp_16SmiTags_e[which(colSums(tp_vp_16SmiTags_e)<167)]

#remove samples with less than 167 reads in the "TP_VP_16S_miTags_99tb_chlp_cyanob" and "TP_VP_16S_miTags_99tb_euk" datasets (so that we can compare the relative abundance of 16S and 18S OTUs considering the same samples)
tp_vp_16SmiTags_e_v2<-subset(tp_vp_16SmiTags_e_min167, select=-c(X123_DCM_0d1.0d2, X125_DCM_0d1.0d2, MP0317, MP0320, MP0322, MP0536, MP0784, MP0882, MP0884, MP0886, MP1164, MP1166, MP1178, MP1337, MP1413, MP1415, MP1419, MP1523, MP1525, MP1599, MP1639, MP1677, MP1678, MP1679, MP1681, MP1682, MP1849, MP1851, MP1853, MP1879, MP2000, MP2043, MP2235, MP2237, MP2658, MP2817, MP427, MP46, MP464, MP468, MP692))
```

Table dimensions and content outline:

```{r starting_dataset, echo=FALSE}
dim(tp_vp_16SmiTags_e_v2)
tp_vp_16SmiTags_e_v2[1:5,1:5]
```

Minimum number of reads per station:

```{r reads_per_sample_overview1, echo=1}
min(colSums(tp_vp_16SmiTags_e_v2)) 
```

Maximum number of reads per station:

```{r reads_per_sample_overview2, echo=1}
max(colSums(tp_vp_16SmiTags_e_v2)) 
```

Identification of station with higher number of reads:

```{r reads_per_sample_overview3, echo=TRUE}
amplicons_per_sample_tp_vp_16SmiTags_e<-colSums(tp_vp_16SmiTags_e_v2)
amplicons_per_sample_tp_vp_16SmiTags_e[which(colSums(tp_vp_16SmiTags_e_v2)>34600)]
```

Overall reads per sample:

``` {r reads_per_sample_overview4, echo=FALSE}
plot(sort(colSums(tp_vp_16SmiTags_e_v2)), pch=19, xlab="sample", ylab="reads per sample", cex=0.9)
```

# 2) Normalization to 167 reads per sample

Let's normalize the original dataset by randomly subsampling 167 reads in each station:

``` {r species_richness_rarefaction1, echo=TRUE}
library(vegan)
tp_vp_16SmiTags_e_v2_t<-t(tp_vp_16SmiTags_e_v2)
tp_16mt_e_ss167<-rrarefy(tp_vp_16SmiTags_e_v2_t, 167)
```

The normalized table shows the following dimensions and format:

```{r species_richness_rarefaction2, echo=FALSE}
dim(tp_16mt_e_ss167)
tp_16mt_e_ss167[1:5,1:5]
```

Its content fits with the expected normalization values (167 reads per station):

``` {r species_richness_rarefaction3, echo=TRUE}
rowSums(tp_16mt_e_ss167)
```

Let's check out how many OTUs don't appear in the new table:

```{r species_richness_rarefaction4, echo=1:5}
length(which(colSums(tp_16mt_e_ss167)==0)) 
```

There are 3854 OTUs that don't show any occurrence in the normalized data. Let's remove them from the table and take a look at its final dimensions:

```{r species_richness_rarefaction5, echo=1:3}
tp_16mt_e_ss167_no_cero<-tp_16mt_e_ss167[,-(which(colSums(tp_16mt_e_ss167)==0))]
dim(tp_16mt_e_ss167_no_cero)

#The final dimensions of the normalized table are  220 3352.
#3352+3854 = 7206
```

Datasets summary:

tp_vp_16SmiTags_e_v2_t --> 220 7206

tp_16mt_e_ss167_no_cero --> 220 3352


# 3) General community analysis

## 3.1) Richness and evenness (Shannon index)

```{r shannon_index1, echo=FALSE}
tp_16mt_e_ss167_div <- diversity(tp_16mt_e_ss167_no_cero, index="shannon")
```

Most of the samples take Shannon Index values between 4.25 and 4.75:

```{r shannon_index2, echo=FALSE}
boxplot(tp_16mt_e_ss167_div, pch=19, main="Shannon's index of diversity")
plot(sort(tp_16mt_e_ss167_div), pch=19, main="Shannon's index of diversity")
```

## 3.2) Richness: OTU number

```{r richness_otu_no1, echo=FALSE}
OTUs_per_sample_tp_16mt_e_ss167<-specnumber(tp_16mt_e_ss167_no_cero)
```

Lowest number of OTUs per sample:

```{r richness_otu_no2, echo=FALSE}
min(OTUs_per_sample_tp_16mt_e_ss167)
```

Maximum number of OTUs per sample:

```{r richness_otu_no3, echo=FALSE}
max(OTUs_per_sample_tp_16mt_e_ss167)
```

In most of the samples, we can identify between 100 and 130 OTUs:

```{r richness_otu_no4, echo=TRUE}
plot(sort(OTUs_per_sample_tp_16mt_e_ss167), pch=19)
boxplot(OTUs_per_sample_tp_16mt_e_ss167, pch=19)
```

## 3.3) Index of evenness

### 3.3.1) Pielou's index

```{r pielou_index_of_evenness1, echo=TRUE}
pielou_evenness_tp_16mt_e_ss167 <- tp_16mt_e_ss167_div/log(OTUs_per_sample_tp_16mt_e_ss167)
```

The Pielou index (constrained between 0 and 1) takes values closer to 1 as the variation of species proportion in a sample decreases. Most of our samples get values between 0.925 and 0.975, meaning that the numerical composition of different OTUs within our samples is so much similar:

```{r pielou_index_of_evenness2, echo=TRUE}
plot(sort(pielou_evenness_tp_16mt_e_ss167), pch=19)
boxplot(pielou_evenness_tp_16mt_e_ss167, pch=19)
```

The OTU_97, with 920 reads, is the most abundant in the overall dataset:

```{r OTUs_overall_abundance, echo=TRUE}
head(sort(colSums(tp_16mt_e_ss167_no_cero), decreasing=T), n=10L)
```

Most of the OTUs show few occurrences, and we probably will be able to identify a certain amount of rare OTUs:

```{r OTUs_overall_abundance2, echo=TRUE}
plot(log(sort(colSums(tp_16mt_e_ss167_no_cero), decreasing=T)), pch=19)
```

<!---
### 3.3.2) Sads
library(sads)
?sads
--->

## 3.4) Abundance Models

<!---
### 3.4.1) Rank-Abundance or Dominance/Diversity Model ("radfit")

To be solved!!

```{r radfit, echo=FALSE}
#?radfit
#otu_tb18_t[1:5,1:5]

tp_vp_16SmiTags_e_v2_t_radfit<-radfit(colSums(tp_vp_16SmiTags_e_v2_t))
plot(tp_vp_16SmiTags_e_v2_t_radfit)
```
--->

### 3.4.2) Preston's Lognormal Model

According to Preston's lognormal model fit into species frequencies groups, we're missing ~2274 species:

```{r preston_model1, echo=T}
tp_16mt_e_ss167_prestonfit<-prestonfit(colSums(tp_vp_16SmiTags_e_v2_t))
plot(tp_16mt_e_ss167_prestonfit, main="Pooled species")

veiledspec(tp_16mt_e_ss167_prestonfit)
```

When computing Preston's lognormal model fit without pooling data into groups, we seem to miss ~1957 species:

```{r preston_model2, echo=4}
tp_16mt_e_ss167_dist_all<-prestondistr(colSums(tp_vp_16SmiTags_e_v2_t))
plot(tp_16mt_e_ss167_prestonfit, main="All malaspina")
lines(tp_16mt_e_ss167_dist_all, line.col="blue3")

veiledspec(tp_16mt_e_ss167_dist_all)
```

<!---
## 3.5) Rarefaction curve

(To be computed)

```{r rarefraction_curve, echo=TRUE}
#?rarecurve

#str(colSums(otu_tb18_t))

#otus_tb18_colsums<-colSums(otu_tb18_t)

#str(otus_tb18_colsums)
#otu_tb18_colsums<-as.matrix(otu_tb18_colsums)

#otu_tb18_colsums<-t(otu_tb18_colsums)

#otu_tb18_colsums[,1:3]

#rarecurve(otu_tb18_colsums, step = 1, 50000, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
#rarecurve(colSums(otu_tb18_t), step = 1, 50000, xlab = "Sample Size", ylab = "OTUs", label = TRUE)
```
--->

## 3.6) Beta diversity

### 3.6.1) Dissimilarity matrix using Bray-Curtis index:

The Bray-Curtis dissimilarity, constrained between 0 (minimum distance) and 1 (highest dissimilarity) allows us to quantify the differences between samples according to the composition and relative abundance of their OTUs. In our dataset, most of the samples pairs take dissimilarity values between 0.85 and 0.95, meaning that their composition is substantially different.

```{r beta_div1, echo=FALSE}
#?vegdist
tp_16mt_e_ss167_no_cero.bray<-vegdist(tp_16mt_e_ss167_no_cero, method="bray")
boxplot(tp_16mt_e_ss167_no_cero.bray, main="Bray-Curtis dissimilarity matrix")
```

### 3.6.2) Hierarchical clustering

There are no evident clusters that we can distinguish according to the species composition and abundance of the samples.

(To be done: assign Longhurst provinces information to each station and check if any of the central clusters is meaningful regarding to the samples' geographical ubication)

```{r beta_div2, echo=FALSE}
#UPGMA
tp_16mt_e_ss167_no_cero.upgma<-hclust(tp_16mt_e_ss167_no_cero.bray, "average")
plot(tp_16mt_e_ss167_no_cero.upgma, cex=.35, main="Samples Hierarchical Clustering")
```

### 3.6.3) Non-metric multidimensional scaling

More of the same... considering the overall distribution, we're not able to identify any informative clustering pattern. 

The stress parameter takes a value below 0.3, meaning that the plot is valid. 

```{r monoNMDS, echo=F}
#NMDS
tp_16mt_e_ss167_no_cero.nmds<-monoMDS(tp_16mt_e_ss167_no_cero.bray)
tp_16mt_e_ss167_no_cero.nmds
plot(tp_16mt_e_ss167_no_cero.nmds, main="monoMDs method")
```

When implementing a most robut function for computing NMDS plots, the result is quiet the same:

```{r metaNMDS, echo=F}
tp_16mt_e_ss167_no_cero.meta_nmds<-metaMDS(tp_16mt_e_ss167_no_cero.bray)
plot(tp_16mt_e_ss167_no_cero.meta_nmds, main="metaMDS method")
```

# 4) Geographical analysis

```{r load_geo_data, echo=F, results="hide", message=F}
#load geographical ubication of stations and sort according to otu_tb16 stations sequence.
TP_geo_16S_e_ss167<-read.table(file="/home/laura/Documents/TFM/home/data/TARASPINA/geo_ubication_SRF_to_1200_220sp.txt", sep="\t", header=T)

row.names(TP_geo_16S_e_ss167)<-TP_geo_16S_e_ss167[,1]
TP_geo_16S_e_ss167_sorted<-TP_geo_16S_e_ss167[row.names(tp_16mt_e_ss167_no_cero),]

dim(TP_geo_16S_e_ss167)
TP_geo_16S_e_ss167[1:5,1:4]
tp_16mt_e_ss167_no_cero[1:5,1:5]

#read lat-long in decimal degrees and translate into distance in km.
library(fossil)

#select only columns containing info about station, latitude and longitude.
TP_geo_16S_e_ss167_v2<-create.lats(TP_geo_16S_e_ss167, loc="sample", long="long", lat="lat")
head(TP_geo_16S_e_ss167)

#create a distance matrix (lower triangle) between a list of points.
geo_distances_TP_16S_e_ss167<-earth.dist(TP_geo_16S_e_ss167_v2, dist = TRUE)
head(geo_distances_TP_16S_e_ss167)
dim(geo_distances_TP_16S_e_ss167)

geo_distances_TP_16S_e_ss167<-as.matrix(geo_distances_TP_16S_e_ss167)
dim(geo_distances_TP_16S_e_ss167)

#geo distances dataset ready to use "geo_distances_MP_euks"
```

Working datasets:

1) Community matrix: tp_16mt_e_ss167_no_cero

```{r working_datasets1, echo=T}
dim(tp_16mt_e_ss167_no_cero)
tp_16mt_e_ss167_no_cero[1:5, 1:5]
```

2) Community Bray-Curtis: otu_tb16_cc_t_ss5000_no_cero.bray

```{r working_datasets2, echo=2}
#dim(otu_tb16_t_ss5000_no_cero.bray)
tp_16mt_e_ss167_no_cero.bray<-as.matrix(tp_16mt_e_ss167_no_cero.bray)
```

3) Stations distances in km: geo_distances_TP_16S_e_ss167

```{r working_datasets3, echo=T}
dim(geo_distances_TP_16S_e_ss167)
```

Communities quickly change their composition across geographical distances:

```{r working_datasets4, echo=T}
plot(geo_distances_TP_16S_e_ss167, tp_16mt_e_ss167_no_cero.bray, pch=19, cex=0.4, xlab="Geopgraphical distances", ylab="Bray-Curtis dissimilarities")
```

## 4.1) Mantel correlograms

When implementing the Mantel Test, we obtain a very low statistic, meaning that the correlation between samples dissimilarity and geographical distances is weak; however, it is significant, so we can consider the test meaningful.

```{r mantel_correlogram1, echo=T}
mantel(geo_distances_TP_16S_e_ss167, tp_16mt_e_ss167_no_cero.bray)
```

Maximum distance between samples:

```{r mantel_correlogram2, echo=F}
max(geo_distances_TP_16S_e_ss167)
```

Minimum distance between samples:

```{r mantel_correlogram3, echo=F}
min(geo_distances_TP_16S_e_ss167)
```

Correlograms:

```{r mantel_correlogram4, echo=T}

TP_16S_e_ss167_mantel_correl_by_1000km<-mantel.correlog(tp_16mt_e_ss167_no_cero.bray, D.geo=geo_distances_TP_16S_e_ss167, break.pts=seq(0,20000, by=1000))
plot(TP_16S_e_ss167_mantel_correl_by_1000km)

TP_16S_e_ss167_mantel_correl_by_100km<-mantel.correlog(tp_16mt_e_ss167_no_cero.bray, D.geo=geo_distances_TP_16S_e_ss167, break.pts=seq(0,20000, by=100))
plot(TP_16S_e_ss167_mantel_correl_by_100km)
```

# 5) Abundance & occurence

```{r OTUs_mean_relative_abund, echo=F, results="hide"}
tp_16mt_e_ss167_no_cero[1:5,1:5]
tp_16mt_e_ss167_no_cero_t<-t(tp_16mt_e_ss167_no_cero)

colSums(tp_16mt_e_ss167_no_cero_t)

#local abundance percentage
tp_16mt_e_ss167_no_cero_t.rabund<-tp_16mt_e_ss167_no_cero_t/167

colSums(tp_16mt_e_ss167_no_cero_t.rabund)
tp_16mt_e_ss167_no_cero_t.rabund[1:5,1:5]

#OTUs mean relative abundance
tp_16mt_e_ss167_no_cero_t.rabund_means<-rowMeans(tp_16mt_e_ss167_no_cero_t.rabund) 
tp_16mt_e_ss167_no_cero_t.rabund_means<-as.data.frame(tp_16mt_e_ss167_no_cero_t.rabund_means)

head(tp_16mt_e_ss167_no_cero_t.rabund_means)
```

```{r OTUs_occurence, echo=F, results='hide'}
tp_16mt_e_ss167_no_cero_t.rabund.occur<-tp_16mt_e_ss167_no_cero_t.rabund
tp_16mt_e_ss167_no_cero_t.rabund.occur[tp_16mt_e_ss167_no_cero_t.rabund.occur>0]<-1
tp_16mt_e_ss167_no_cero_t.rabund.occur[1:5,1:5] ### presence - absence table

#percentage of occurence in overall stations
tp_16mt_e_ss167_no_cero_t.rabund_means.occurence_perc<-as.data.frame(100*(rowSums(tp_16mt_e_ss167_no_cero_t.rabund.occur)/220))

str(tp_16mt_e_ss167_no_cero_t.rabund_means.occurence_perc)
```

```{r merge_rabund_peroccur, echo=F, results='hide'}
tp_16mt_e_ss167_rabund_percoccur<-merge(tp_16mt_e_ss167_no_cero_t.rabund_means, tp_16mt_e_ss167_no_cero_t.rabund_means.occurence_perc, by="row.names")

colnames(tp_16mt_e_ss167_rabund_percoccur)<-c("OTUs","mean_rabund","perc_occur")
tp_16mt_e_ss167_rabund_percoccur[1:5,]

row.names(tp_16mt_e_ss167_rabund_percoccur)<-tp_16mt_e_ss167_rabund_percoccur[,1]
tp_16mt_e_ss167_rabund_percoccur<-tp_16mt_e_ss167_rabund_percoccur[,-1]
tp_16mt_e_ss167_rabund_percoccur[1:5,]
```

In the following plot, we can appreciate the OTUs distribution according to their percentage of occurence and relative abundance. The blue line restricts regionally abundant OTUs (> 0.1%).

```{r abund_vs_occurence_table, echo=F}
plot(tp_16mt_e_ss167_rabund_percoccur$mean_rabund,tp_16mt_e_ss167_rabund_percoccur$perc_occur, log="x", pch=19, cex=0.8, xlab="Mean relative abundance", ylab="Percentage of occurence")
abline(h=80, col="red") #occurence higher than 80%
abline(v=0.00001, col="green") #rare OTUs
abline(v=0.001, col="blue") #cosmopolitan OTUs

#Conventional limits:
#Regionally rare     = 0.00001
#Regionally abundant = 0.001
```

Regionally abundant OTUs (relative abundance over 0.1%):

```{r abundant_OTUs, echo=7}
#regionally abundant
tp_16mt_e_ss167_abundant<-tp_16mt_e_ss167_rabund_percoccur[tp_16mt_e_ss167_rabund_percoccur$mean_rabund > 0.001,]

tp_16mt_e_ss167_abundant_sorted<-tp_16mt_e_ss167_abundant[order(tp_16mt_e_ss167_abundant$mean_rabund, tp_16mt_e_ss167_abundant$perc_occur, decreasing = T), c(1,2)]

tp_16mt_e_ss167_abundant_sorted
dim(tp_16mt_e_ss167_abundant_sorted)
```

Proportion of regionally abundant OTUs (%):

```{r abundant_OTUs2, echo=F}
#there are 84 regionally abundant OTUs.
(208/3344)*100 # = 6.22% of the OTUs are regionally abundant

#length(row.names(otu_tb16_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund > 0.001,])) # 83 OTUs
#row.names(otu_tb16_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund > 0.001,])
```

Cosmopolitan OTUs (abundance over 0.1% and occurence in more than 80% of samples):

```{r select_cosmopolitan, echo=6}
tp_16mt_e_ss167_rabund_cosm<-tp_16mt_e_ss167_rabund_percoccur[tp_16mt_e_ss167_rabund_percoccur$mean_rabund > 0.001,]
tp_16mt_e_ss167_rabund_poccur_cosm<-tp_16mt_e_ss167_rabund_cosm[tp_16mt_e_ss167_rabund_cosm$perc_occur > 80,]
tp_16mt_e_ss167_cosmop_sorted<-tp_16mt_e_ss167_rabund_poccur_cosm[order(tp_16mt_e_ss167_rabund_poccur_cosm$perc_occur, tp_16mt_e_ss167_rabund_poccur_cosm$mean_rabund, decreasing = T), c(1,2)]

tp_16mt_e_ss167_cosmop_sorted
dim(tp_16mt_e_ss167_cosmop_sorted)
```

Proportion of cosmopolitan OTUs (%):

```{r percentage_cosmopolitan, echo=F}
(0/3344)*100
```

Number and proportion (%) of rare OTUs:

```{r rare_OTUs, echo=1}
dim(tp_16mt_e_ss167_rabund_percoccur[tp_16mt_e_ss167_rabund_percoccur$mean_rabund < 0.00001 & tp_16mt_e_ss167_rabund_percoccur$mean_rabund >0,])
 
(0/3344)*100 # = 29.22% of the OTUs are regionally rare
```

<!---
```{r otu_col_chech, echo = T}
dim(tp_16mt_e_ss167_rabund_percoccur)

#rare OTUs:
# length(row.names(otu_tb16_cc_ss5000_rabund_percoccur[otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund < 0.00001 & otu_tb16_cc_ss5000_rabund_percoccur$mean_rabund >0 ,])) # 325 OTUs
```
--->